---
title: "Differential Expression Analysis"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: readable
    highlight: zenburn
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message = FALSE, warning = FALSE}
library(DESeq2)
library(EnhancedVolcano)
library(tidyverse)
library(cowplot)
library(pheatmap)
library(viridis)
```

## 1. Import files

We load the normalized `dds` object created during preprocessing and batch correction.
```{r load_data}
load(file = "../results/dds_norm.RData")
dds_norm
```

We load the metadata containing project and experimental condition information.
```{r load_metadata}
metadatos <- read.csv(file = "../data/metadata_all.csv", row.names = 1)
head(metadatos)
```

---

## 2. Differential Expression Analysis

We check the conditions available in our metadata to set up the comparison.
```{r check_conditions}
metadatos %>%
  group_by(Condition) %>%
  summarise(Count = n())
```

We create the DESeq2 results object, explicitly defining the comparison: Control vs Treatment.
*Note: in DESeq2 the standard contrast format is `c("factor", "numerator", "denominator")`.*
```{r deseq_results}
res <- results(dds_norm, contrast = c("Condition", "Control", "Treatment"))
res

write.csv(res, file = "../results/deseq_results_Control_vs_Treatment.csv")
```

### 2.1. Volcano Plot
We create a volcano plot for a general view of the differential expression. We set the significance threshold (`padj < 0.01`) and the fold change threshold (`|log2FC| > 1`).

```{r volcano_plot, fig.width=11, fig.height=8.5, warning=FALSE}
volc <- EnhancedVolcano(res,
               lab = rownames(res),
               x = 'log2FoldChange',
               y = 'padj',
               axisLabSize = 12,
               pCutoff = 0.01,
               FCcutoff = 1, # Matches the filtering step below
               col = c('grey', 'orange', 'purple', 'forestgreen'),
               colAlpha = 0.7,
               pointSize = 3,
               title = 'Volcano Plot',
               titleLabSize = 12,
               subtitle = 'Control vs Treatment',
               subtitleLabSize = 10,
               captionLabSize = 10,
               labSize = 3,
               legendLabels = c('NS', '|log2FC|>1', 'adj-p<0.01', 'DEG'),
               legendPosition = 'right',
               legendLabSize = 12,
               border = 'full')
volc

ggsave(filename = "../results/volcano_Control_vs_Treatment.pdf",
       plot = volc, width = 11, height = 8.5)
```

### 2.2. Extract Differentially Expressed Genes (DEGs)
We filter the results to keep only genes that pass the significance and fold change thresholds.

```{r filter_degs}
deg <- subset(res, padj < 0.01 & abs(log2FoldChange) > 1)
paste("Total DEGs found:", nrow(deg))

up <- subset(deg, log2FoldChange > 1)
paste("Up-regulated genes:", nrow(up))

down <- subset(deg, log2FoldChange < (-1))
paste("Down-regulated genes:", nrow(down))

write.csv(deg, file = "../results/deg_Control_vs_Treatment.csv", row.names = TRUE)
```

A sample output of this step is provided in `data_example/deg_Control_vs_Treatment.csv` so the next steps can be tested without re-running the full pipeline.

### 2.3. Expression Heatmap
We generate a heatmap to visualize expression patterns of the identified DEGs across samples.

```{r prepare_heatmap_data}
normcounts <- counts(dds_norm, normalized = TRUE)
colnames(normcounts) <- metadatos$Replicates

muestras_contraste <- metadatos %>%
  filter(Condition %in% c("Control", "Treatment")) %>%
  pull(Replicates)

subnormcounts <- normcounts[, muestras_contraste]
deg_id <- rownames(deg)
```

```{r plot_heatmap, fig.width=11, fig.height=8.5}
hm <- pheatmap(subnormcounts[deg_id, ],
         scale = "row",
         show_rownames = FALSE,
         treeheight_row = 0,
         color = viridis(100),
         main = "Heatmap of DEGs: Control vs Treatment")
hm

ggsave(filename = "../results/heatmap_Control_vs_Treatment.pdf",
       plot = hm, width = 11, height = 8.5)
```

---

## 3. Consolidate DEG lists

We save the DEG IDs in a clean table format for downstream functional analysis.

*(This section is ready to scale up if you add comparisons from other experiments later.)*

```{r consolidate_degs}
BP1_deg <- read.csv(file = "../results/deg_Control_vs_Treatment.csv", row.names = 1)

deg_summary <- data.frame(
  GeneID = rownames(BP1_deg),
  Contrast = "Control_vs_Treatment"
)

head(deg_summary)

write_csv(deg_summary, file = "../results/common_unique_degs.csv")
```

---

## Session information
```{r session_info}
sessionInfo()
```
