---
title: "Exploratory Analysis"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: zenburn
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message = FALSE, warning = FALSE}
library(cowplot)
library(viridis)
library(tximport)
library(DESeq2)
library(tidyverse)
```

> **Before running this script:** confirm your samples passed the taxonomic check in `01_taxonomic_assessment/`. This script assumes Kallisto quantification (`00_preprocessing_linux/05_quantification.sh`) is already complete.

## 1. Import files and preprocess data

### 1.1. Import count files (Kallisto)
We set the path and pattern for the Kallisto `abundance.tsv` files to load the counts.

```{r import_counts}
files <- list.files(full.names = TRUE,
                    path = "../data/counts",
                    pattern = "*.tsv")
files
```

### 1.2. Import experimental metadata
We load the metadata, which contains the conditions and replicates for our experiment.

```{r import_metadata}
metadatos <- read.csv(file = "../data/metadata_all.csv", row.names = 1)
head(metadatos)
```

**Quality check:** we verify if the number of count files matches the number of samples in our metadata.
```{r check_match}
length(files) == nrow(metadatos)
```

### 1.3. Build the tx2gene object
Because we are working with a **bacterial genome**, there is no alternative splicing. The transcript name is therefore the same as the gene name, so `TXNAME` and `GENEID` map 1:1.

*Note: we read the first count file to get the gene IDs without hardcoding any local paths.*

```{r build_tx2gene}
kallisto_ejemplo <- read.delim(files[1])

tx2gene <- data.frame(
  TXNAME = kallisto_ejemplo$target_id,
  GENEID = kallisto_ejemplo$target_id
)
```

### 1.4. Import with tximport and create the DESeq2 object
We merge the counts and set up the `DESeqDataSet` object, using `Condition` as our main experimental variable.

```{r run_tximport}
txi.kallisto <- tximport(files,
                         type = "kallisto",
                         tx2gene = tx2gene)

dds <- DESeqDataSetFromTximport(txi.kallisto,
                                colData = metadatos,
                                design = ~ Condition)
dds
```

### 1.5. Extract and filter counts
To improve statistical power and reduce noise, we remove genes with low expression. We keep genes that have an average of more than 10 counts across all samples — adjust this threshold based on your sequencing depth and number of replicates.

```{r raw_and_filtered_counts}
rawcounts <- as.matrix(counts(dds))
head(rawcounts)
write.csv(rawcounts, file = "../results/rawcounts.csv", row.names = TRUE)

filtcounts <- rawcounts[rowMeans(rawcounts) > 10, ]
head(filtcounts)

p <- (nrow(filtcounts) / nrow(rawcounts)) * 100
paste("Expressed genes kept after filtering =", round(p, 2), "%")

write.csv(filtcounts, file = "../results/filtcounts.csv", row.names = TRUE)
```

We update our `DESeqDataSet` object to only use the genes we kept.

```{r update_dds}
expgen <- rownames(filtcounts)
dds <- dds[expgen, ]
dds
```

---

## 2. Exploratory data analysis

### 2.1. Variance Stabilizing Transformation (VST)
We apply the VST to normalize the data. This makes the variance independent of the mean, which helps clustering methods like PCA work better.

```{r vst_transformation}
vsd <- vst(dds, blind = FALSE)
```

### 2.2. Principal Component Analysis (PCA)
We calculate and plot the PCA to see how similar the samples are and check if they group by their experimental condition.

```{r pca_calculation}
pcaData <- plotPCA(vsd, intgroup = "Condition", returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

paste("Total explained variance (PC1 + PC2) =", sum(percentVar), "%")
```

Create and save the PCA plot:

```{r pca_plot, fig.width=11, fig.height=8.5}
pca <- ggplot(pcaData,
              aes(x = PC1,
                  y = PC2,
                  color = Condition,
                  shape = Condition)) +
  geom_point(size = 7, alpha = 0.8) +
  scale_shape_manual(values = 15:18) +
  geom_label(aes(label = name), show.legend = FALSE) +
  labs(title = "PCA: Treatment vs Control",
       subtitle = "Filtered and VST normalized data",
       x = paste0("PC1: ", percentVar[1], "% variance"),
       y = paste0("PC2: ", percentVar[2], "% variance"),
       color = "Condition",
       shape = "Condition") +
  theme_minimal() +
  coord_fixed()

pca

ggsave(filename = "../results/pca_raw.pdf", plot = pca, width = 11, height = 8.5)
```

### 2.3. Save workspace
We save the objects in an `.RData` file so subsequent scripts can reuse them without re-running this analysis.

```{r save_rdata}
save(metadatos, dds, rawcounts, filtcounts, pca, file = "../results/dds.RData")
```

---

## Session information
Recording the R version and packages used helps make sure this analysis can be reproduced.

```{r session_info}
sessionInfo()
```
