---
title: "Advanced Analysis: GSEA, Trait-Specific Profiling, and Networks"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: zenburn
  pdf_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message = FALSE, warning = FALSE}
library(clusterProfiler) # Enrichment analysis (GO/KEGG)
library(enrichplot)      # Plot enrichment results
library(tidyverse)       # Data organizing and cleaning
library(ggplot2)         # Custom plots
library(pheatmap)        # Heatmaps
library(viridis)         # Color palettes
library(ggrepel)         # Avoid text overlap in plots
library(KEGGREST)        # Pathway names from the KEGG API
library(STRINGdb)        # Protein-protein interactions
library(igraph)          # Build and analyze networks
library(ggraph)          # Graph networks
library(tidygraph)       # Modify networks
```

## 1. Environment Setup

```{r create_directories}
adv_dir <- "../results/advanced_analysis"
dir.create(adv_dir, showWarnings = FALSE, recursive = TRUE)
```

## 2. Data Import

We load the ranked list (ordered by log2FC) and the full annotated results. GSEA requires the complete gene list, not just the significant subset.

```{r load_data}
ranked_df <- readRDS("../results/advanced_analysis/gsea_ranked_list.rds") %>%
  group_by(GeneID) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

gsea_vector <- ranked_df$log2FoldChange
names(gsea_vector) <- ranked_df$GeneID
gsea_vector <- sort(gsea_vector, decreasing = TRUE)

full_annotation <- read_csv("../results/advanced_analysis/full_annotated_results.csv")

eggnog_data <- read.delim("../data/annotation.tsv")

term2gene_kegg <- eggnog_data %>%
  dplyr::select(KEGG_Pathway, X.query) %>%
  tidyr::separate_rows(KEGG_Pathway, sep = ",") %>%
  dplyr::filter(grepl("map", KEGG_Pathway)) %>%
  dplyr::mutate(query = trimws(X.query), KEGG_Pathway = trimws(KEGG_Pathway)) %>%
  dplyr::select(KEGG_Pathway, X.query) %>%
  as.data.frame()
```

---

## 3. Gene Set Enrichment Analysis (GSEA)

GSEA looks at the whole ranked gene list to detect pathways that are globally up- or down-regulated, even when individual genes don't pass strict significance cutoffs.

```{r run_gsea, warning=FALSE, message=FALSE}
kegg_names <- keggList("pathway")

term2name_kegg <- data.frame(
  PathwayID = sub("path:", "", names(kegg_names)),
  PathwayName = as.character(kegg_names)
)
term2name_kegg$PathwayName <- gsub(" - Reference pathway", "", term2name_kegg$PathwayName)

gsea_kegg <- GSEA(geneList = gsea_vector,
                  TERM2GENE = term2gene_kegg,
                  TERM2NAME = term2name_kegg,
                  pvalueCutoff = 0.05,
                  pAdjustMethod = "BH",
                  minGSSize = 10,
                  maxGSSize = 500,
                  eps = 0)
```

### 3.1. Systemic Shifts (Ridgeplot)

```{r ridgeplot, fig.width=10, fig.height=8}
p_ridge <- ridgeplot(gsea_kegg, showCategory = 10) +
  labs(title = "GSEA Ridgeplot: Systemic KEGG Pathway Shifts",
       x = "Enrichment Distribution (Log2FC)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16))

p_ridge
ggsave(filename = paste0(adv_dir, "/gsea_ridgeplot.pdf"), plot = p_ridge, width = 10, height = 8)
```

### 3.2. Top Enriched Pathway Plot

```{r gsea_plot, fig.width=10, fig.height=8.5}
# Index 2 is used here to skip generic housekeeping pathways (e.g., Ribosome)
# that often rank first but add little biological insight. Adjust as needed.
target_pathway_idx <- 2
top_gsea_pathway <- gsea_kegg@result$ID[target_pathway_idx]

p_gseaplot <- gseaplot2(gsea_kegg,
                        geneSetID = top_gsea_pathway,
                        title = paste("GSEA - Pathway:", gsea_kegg@result$Description[target_pathway_idx]),
                        pvalue_table = TRUE)

pdf(paste0(adv_dir, "/target_pathway_gseaplot.pdf"), width = 10, height = 8.5)
print(p_gseaplot)
dev.off()
```

---

## 4. Trait-Specific Gene Profiling

This section filters DEGs by keyword to highlight genes relevant to a specific biological question — for example, virulence factors in a pathogen, stress-response genes, or any other functional category of interest. **Adjust the keyword list below to match your research question.**

```{r extract_trait_genes}
# Example: virulence-associated keywords for a plant-pathogenic bacterium.
# Replace with keywords relevant to your organism and hypothesis.
trait_keywords <- "pectate lyase|cellulase|protease|secretion system|pel|peh|out|kdg"

trait_df <- full_annotation %>%
  filter(padj < 0.05) %>%
  filter(str_detect(tolower(Description), trait_keywords) |
         str_detect(tolower(Preferred_name), trait_keywords)) %>%
  arrange(desc(abs(log2FoldChange)))

trait_genes <- trait_df$GeneID
message("Found ", length(trait_genes), " genes matching the trait keywords.")

write_csv(trait_df, paste0(adv_dir, "/trait_genes_table.csv"))
```

### 4.1. Trait-Specific Heatmap

```{r trait_heatmap, fig.width=8, fig.height=10}
load(file = "../results/dds_norm.RData")
metadatos <- read.csv(file = "../data/metadata_all.csv", row.names = 1)

colnames(normcounts) <- metadatos$Replicates

rownames(normcounts) <- str_extract(rownames(normcounts), "WP_[0-9]+\\.[0-9]+")
normcounts <- normcounts[!is.na(rownames(normcounts)), ]

trait_matrix <- normcounts[trait_genes, ]

rownames(trait_matrix) <- ifelse(!is.na(trait_df$Preferred_name),
                                 trait_df$Preferred_name,
                                 trait_df$GeneID)

p_trait_hm <- pheatmap(trait_matrix,
                       scale = "row",
                       color = viridis(100),
                       treeheight_row = 15,
                       treeheight_col = 15,
                       main = "Trait-Specific Gene Expression",
                       fontsize_row = 8)

ggsave(filename = paste0(adv_dir, "/heatmap_trait_genes.pdf"), plot = p_trait_hm, width = 8, height = 10)
```

---

## 5. Upstream Regulators Analysis

We search for transcription factors and DNA-binding proteins that might be driving the observed changes.

```{r tf_volcano, fig.width=10, fig.height=8, warning=FALSE}
tf_keywords <- "transcription factor|regulator|DNA-binding"

volcano_data <- full_annotation %>%
  mutate(
    Is_Regulator = if_else(str_detect(tolower(Description), tf_keywords) |
                           str_detect(tolower(Preferred_name), tf_keywords), "Yes", "No"),
    Label = if_else(Is_Regulator == "Yes" & padj < 0.05 & abs(log2FoldChange) > 1,
                    coalesce(Preferred_name, GeneID), "")
  )

p_tf_volcano <- ggplot(volcano_data, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = Is_Regulator, size = Is_Regulator, alpha = Is_Regulator)) +
  scale_color_manual(values = c("No" = "grey80", "Yes" = "firebrick")) +
  scale_size_manual(values = c("No" = 1.5, "Yes" = 3)) +
  scale_alpha_manual(values = c("No" = 0.4, "Yes" = 0.9)) +
  geom_text_repel(aes(label = Label),
                  size = 3.5,
                  box.padding = 0.5,
                  max.overlaps = 20,
                  color = "black",
                  fontface = "bold") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", alpha = 0.5) +
  labs(title = "Global Transcriptional Regulators",
       subtitle = "Highlighted targets are potential control factors",
       x = "Log2 Fold Change",
       y = "-Log10(p-adj)") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold", size = 16))

p_tf_volcano
ggsave(filename = paste0(adv_dir, "/volcano_regulators.pdf"), plot = p_tf_volcano, width = 10, height = 8)
```

---

## 6. Secondary Trait Profiling (Example: Bioenergetics)

A second example of keyword-based gene filtering, this time for energy metabolism and ion transport genes. This section demonstrates how the pattern in Section 4 can be reused for any trait of interest.

```{r secondary_trait_targets}
secondary_keywords <- "atp synthase|efflux|pump|ion channel|cytochrome|electron transport|nadh dehydrogenase|proton|potassium|sodium|calcium|voltage"

secondary_genes <- full_annotation %>%
  filter(padj < 0.05) %>%
  filter(str_detect(tolower(Description), secondary_keywords) |
         str_detect(tolower(Preferred_name), secondary_keywords) |
         str_detect(tolower(GOs), "ion transport|atp biosynthetic")) %>%
  arrange(desc(log2FoldChange))

message("Found ", nrow(secondary_genes), " genes matching secondary trait keywords.")

write_csv(secondary_genes, paste0(adv_dir, "/secondary_trait_targets.csv"))
```

---

## 7. Network Analysis (PPI Hubs & Clusters)

We build a Protein-Protein Interaction (PPI) network using STRING's mapping tool. A high-confidence threshold (score > 700) removes noise and isolates distinct functional modules.

```{r ppi_topology, warning=FALSE, message=FALSE}
deg_network <- full_annotation %>%
  filter(padj < 0.05) %>%
  dplyr::select(GeneID, Preferred_name, seed_ortholog, log2FoldChange) %>%
  mutate(search_name = coalesce(Preferred_name, GeneID)) %>%
  group_by(search_name) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1, with_ties = FALSE) %>%
  ungroup()

valid_orthologs <- deg_network$seed_ortholog[!is.na(deg_network$seed_ortholog) & deg_network$seed_ortholog != ""]
tax_id <- as.numeric(strsplit(valid_orthologs[1], "\\.")[[1]][1])

string_db <- STRINGdb$new(version = "11.5", species = tax_id, score_threshold = 700)

mapped_genes <- string_db$map(as.data.frame(deg_network), "search_name", removeUnmappedRows = TRUE)
interactions <- string_db$get_interactions(mapped_genes$STRING_id)

ppi_graph <- graph_from_data_frame(d = interactions, directed = FALSE)
ppi_graph <- simplify(ppi_graph)

if (vcount(ppi_graph) > 0) {
  communities <- cluster_louvain(ppi_graph)
  ppi_graph <- set_vertex_attr(ppi_graph, "cluster", value = as.numeric(membership(communities)))

  ppi_graph <- set_vertex_attr(ppi_graph, "degree", value = degree(ppi_graph))
  ppi_graph <- set_vertex_attr(ppi_graph, "betweenness", value = betweenness(ppi_graph))

  node_metadata <- data.frame(STRING_id = V(ppi_graph)$name) %>%
    left_join(mapped_genes, by = "STRING_id") %>%
    distinct(STRING_id, .keep_all = TRUE)

  ppi_graph <- set_vertex_attr(ppi_graph, "label", value = node_metadata$search_name)
  ppi_graph <- set_vertex_attr(ppi_graph, "log2FC", value = node_metadata$log2FoldChange)

  # Hubs are selected by betweenness (bottlenecks), which highlights genes that
  # bridge distinct functional modules rather than just locally well-connected ones.
  top_hubs <- node_metadata %>%
    mutate(Degree = V(ppi_graph)$degree,
           Betweenness = V(ppi_graph)$betweenness,
           Cluster = V(ppi_graph)$cluster) %>%
    group_by(Cluster) %>%
    slice_max(order_by = Betweenness, n = 3, with_ties = FALSE) %>%
    ungroup() %>%
    arrange(desc(Betweenness)) %>%
    head(15)

  write_csv(top_hubs, paste0(adv_dir, "/ppi_bottleneck_hubs.csv"))
} else {
  message("STRINGdb could not find any interactions above the 700 threshold.")
}
```

### 7.1. Network Visualization (ggraph)

We use `ggraph` with a Kamada-Kawai layout to separate distinct functional modules, while `ggrepel` highlights the bottleneck hubs identified above.

```{r plot_network, fig.width=12, fig.height=12, warning=FALSE}
library(ggraph)
library(tidygraph)

if (vcount(ppi_graph) > 0) {
  net_components <- components(ppi_graph)

  if (max(net_components$csize) > 1) {
    main_component <- induced_subgraph(ppi_graph, vids = which(net_components$membership == which.max(net_components$csize)))

    tidy_net <- as_tbl_graph(main_component) %>%
      mutate(
        is_hub = name %in% top_hubs$STRING_id,
        hub_label = ifelse(is_hub, label, NA),
        clean_betweenness = replace_na(betweenness, 0),
        max_bet = max(clean_betweenness, na.rm = TRUE),
        node_size = ifelse(max_bet > 0, (clean_betweenness / max_bet) * 10 + 3, 5)
      )

    p_net <- ggraph(tidy_net, layout = "kk") +
      geom_edge_link(color = "grey85", width = 0.5, alpha = 0.6) +
      geom_node_point(aes(size = node_size, color = as.factor(cluster), alpha = is_hub)) +
      scale_alpha_manual(values = c("TRUE" = 1, "FALSE" = 0.4)) +
      geom_node_text(aes(label = hub_label),
                     repel = TRUE,
                     fontface = "bold",
                     size = 5,
                     box.padding = 1,
                     point.padding = 0.5,
                     max.overlaps = Inf,
                     bg.color = "white",
                     bg.r = 0.15,
                     segment.color = "grey40",
                     segment.size = 0.6) +
      scale_color_viridis_d(option = "viridis") +
      labs(title = "Protein-Protein Interaction Network",
           subtitle = "Top 15 Bottleneck Hubs Highlighted (High Confidence STRING Score > 700)") +
      theme_void() +
      theme(legend.position = "none",
            plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
            plot.subtitle = element_text(size = 14, hjust = 0.5, color = "grey40"))

    pdf(paste0(adv_dir, "/ppi_high_confidence_network.pdf"), width = 12, height = 12)
    print(p_net)
    dev.off()

    print(p_net)

  } else {
    message("No connected components found to draw the plot.")
  }
} else {
  message("Plot cancelled: the network is empty.")
}
```

---

## Session Information
```{r session_info}
sessionInfo()
```
