---
title: "Functional Enrichment Analysis of DEGs"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: zenburn
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message = FALSE, warning = FALSE}
library(clusterProfiler)
library(enrichplot)
library(dplyr)
library(tidyr)
library(ggplot2)
library(pathview)
library(GO.db)
library(readr)
library(stringr)
library(UpSetR)
library(STRINGdb)
library(magick)
```

> **Annotation requirement:** this script expects an [eggNOG-mapper](http://eggnog-mapper.embl.de/) annotation file (`annotation.tsv`) for your organism. eggNOG-mapper provides GO terms, KEGG pathways, and ortholog assignments directly from protein sequences — useful for bacteria without a curated organism-specific Bioconductor annotation package.

## 1. Setup environment and directories

```{r create_directories}
out_dir <- "../results/enrichment"
kegg_dir <- "../results/enrichment/kegg"
adv_dir <- "../results/advanced_analysis"

dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(kegg_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(adv_dir, showWarnings = FALSE, recursive = TRUE)
```

## 2. Import and prepare data

### 2.1. Load functional metadata (eggNOG)
We load the eggNOG annotation and build `TERM2GENE` objects to map genes to their functions.

```{r load_eggnog}
eggnog_data <- read.delim("../data/annotation.tsv")

# GO Dictionary
term2gene_go <- eggnog_data %>%
  dplyr::select(GOs, X.query) %>%
  tidyr::separate_rows(GOs, sep = ",") %>%
  dplyr::filter(GOs != "" & GOs != "-" & !is.na(GOs)) %>%
  dplyr::mutate(query = trimws(X.query), GOs = trimws(GOs)) %>%
  dplyr::select(GOs, query) %>%
  dplyr::rename(X.query = query) %>%
  as.data.frame()

# Map GO IDs to descriptive terms
vector_go <- Term(GOTERM)
term2name_go <- data.frame(
  GO_ID = as.character(names(vector_go)),
  Term = as.character(unname(vector_go)),
  stringsAsFactors = FALSE
)

# KEGG Dictionary (filtered for 'map' metabolic pathways)
term2gene_kegg <- eggnog_data %>%
  dplyr::select(KEGG_Pathway, X.query) %>%
  tidyr::separate_rows(KEGG_Pathway, sep = ",") %>%
  dplyr::filter(KEGG_Pathway != "" & KEGG_Pathway != "-" & !is.na(KEGG_Pathway)) %>%
  dplyr::filter(grepl("map", KEGG_Pathway)) %>%
  dplyr::mutate(query = trimws(X.query), KEGG_Pathway = trimws(KEGG_Pathway)) %>%
  dplyr::select(KEGG_Pathway, query) %>%
  dplyr::rename(X.query = query) %>%
  as.data.frame()
```

### 2.2. Import Differentially Expressed Genes (DEGs)
We extract and separate the significant genes found in the previous step.

*Note: the regex `WP_[0-9]+\\.[0-9]+` matches NCBI RefSeq protein IDs, the standard format for bacterial gene identifiers. Adjust this pattern if your reference uses a different ID scheme (e.g., locus tags like `b0001`).*

```{r load_degs, warning=FALSE}
deg_table <- read_csv("../results/deg_Control_vs_Treatment.csv") %>%
  mutate(GeneID = str_extract(...1, "WP_[0-9]+\\.[0-9]+"))

up_genes <- deg_table %>% filter(log2FoldChange > 0) %>% pull(GeneID) %>% trimws()
down_genes <- deg_table %>% filter(log2FoldChange < 0) %>% pull(GeneID) %>% trimws()
all_deg_ids <- trimws(deg_table$GeneID)
```

---

## 3. Enrichment Analysis (ORA)

### 3.1. Gene Ontology (GO)
We test for over-represented biological processes in both up- and down-regulated genes.

```{r go_enrichment}
enrich_go_up <- enricher(gene = up_genes,
                         TERM2GENE = term2gene_go,
                         TERM2NAME = term2name_go,
                         pAdjustMethod = "BH",
                         pvalueCutoff  = 0.1,
                         qvalueCutoff  = 0.1)

enrich_go_down <- enricher(gene = down_genes,
                           TERM2GENE = term2gene_go,
                           TERM2NAME = term2name_go,
                           pAdjustMethod = "BH",
                           pvalueCutoff  = 0.1,
                           qvalueCutoff  = 0.1)
```

```{r go_dotplots, fig.width=10, fig.height=7}
p_go_up <- enrichplot::dotplot(enrich_go_up, showCategory = 15, title = "GO Enrichment (Up-regulated)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16), axis.text.y = element_text(size = 11))

p_go_down <- enrichplot::dotplot(enrich_go_down, showCategory = 15, title = "GO Enrichment (Down-regulated)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16), axis.text.y = element_text(size = 11))

p_go_up
p_go_down

ggsave(filename = paste0(out_dir, "/dotplot_GO_UP.pdf"), plot = p_go_up, width = 10, height = 7)
ggsave(filename = paste0(out_dir, "/dotplot_GO_DOWN.pdf"), plot = p_go_down, width = 10, height = 7)
```

### 3.2. KEGG Pathway Enrichment
We check for affected metabolic pathways.

```{r kegg_enrichment, fig.width=10, fig.height=7}
enrich_kegg <- enricher(gene = all_deg_ids,
                        TERM2GENE = term2gene_kegg,
                        pAdjustMethod = "BH",
                        pvalueCutoff  = 0.1,
                        qvalueCutoff  = 0.1)

p_kegg <- enrichplot::dotplot(enrich_kegg, showCategory = 15, title = "KEGG Pathway Enrichment") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16), axis.text.y = element_text(size = 11))

p_kegg
ggsave(filename = paste0(out_dir, "/dotplot_KEGG.pdf"), plot = p_kegg, width = 10, height = 7)
```

---

## 4. Automated KEGG Pathway Visualization (Pathview)

We download the top 15 affected metabolic pathways and color them using `log2FoldChange` values.

```{r prepare_pathview}
eggnog_ko <- eggnog_data %>%
  dplyr::select(X.query, KEGG_ko) %>%
  dplyr::filter(KEGG_ko != "" & KEGG_ko != "-") %>%
  dplyr::mutate(KEGG_ko = gsub("ko:", "", KEGG_ko)) %>%
  dplyr::mutate(KEGG_ko = sapply(strsplit(as.character(KEGG_ko), ","), `[`, 1))

pathview_data <- deg_table %>%
  dplyr::inner_join(eggnog_ko, by = c("GeneID" = "X.query")) %>%
  dplyr::filter(!is.na(KEGG_ko) & !is.na(log2FoldChange))

fc_vector <- pathview_data$log2FoldChange
names(fc_vector) <- pathview_data$KEGG_ko
```

```{r render_pathview_loop, results='hide', message=FALSE, warning=FALSE}
top_kegg_ids <- head(enrich_kegg@result$ID, 15)
top_kegg_clean <- gsub("map", "", top_kegg_ids)

current_wd <- getwd()
setwd(kegg_dir)

# 1. Loop to download and render each map (saved as PNGs)
for (path_id in top_kegg_clean) {
  message("Processing KEGG pathway: ", path_id)

  tryCatch({
    pathview(gene.data  = fc_vector,
             pathway.id = path_id,
             species    = "ko",
             limit      = list(gene = max(abs(fc_vector)), cpd = 1),
             kegg.native = TRUE)
  }, error = function(e) {
    message("Skipped pathway ", path_id, " (insufficient data or API constraint).")
  })
}

# 2. Combine all maps into a single PDF safely
png_files <- list.files(pattern = "\\.pathview\\.png$")

# Safety filter: keep only files larger than 5KB to avoid "ImproperImageHeader"
# errors from failed KEGG downloads (usually 0 bytes or tiny HTML error pages).
valid_png_files <- png_files[file.info(png_files)$size > 5000]

if (length(valid_png_files) > 0) {
  message("Combining ", length(valid_png_files), " valid maps into a single PDF...")

  kegg_maps <- image_read(valid_png_files)
  image_write(kegg_maps, format = "pdf", path = "Top15_KEGG_Pathways_Consolidated.pdf")

  file.remove(png_files)
} else {
  message("No valid maps were generated to combine.")
  file.remove(png_files)
}

setwd(current_wd)
```

---

## 5. Network Analysis (Cnetplot and STRING)

### 5.1. GO Concept Network (Cnetplot)
We visualize how genes connect to their functional terms.

```{r cnetplot, fig.width=12, fig.height=10}
color_vector <- pathview_data$log2FoldChange
names(color_vector) <- trimws(pathview_data$GeneID)

p_cnet <- cnetplot(enrich_go_up,
                   showCategory = 3,
                   foldChange = color_vector) +
  ggtitle("Gene-Concept Network (Up-regulated)") +
  theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))

p_cnet
ggsave(filename = paste0(out_dir, "/cnetplot_GO.pdf"), plot = p_cnet, width = 12, height = 10)
```

### 5.2. Protein-Protein Interactions (PPI - STRINGdb)
We search for functional networks using orthologs identified by eggNOG.

```{r string_network, warning=FALSE}
ppi_data <- deg_table %>%
  dplyr::inner_join(eggnog_data, by = c("GeneID" = "X.query")) %>%
  dplyr::filter(padj < 0.05 & seed_ortholog != "") %>%
  dplyr::arrange(padj) %>%
  dplyr::slice_head(n = 50)

string_ids <- ppi_data$seed_ortholog
tax_id <- as.numeric(strsplit(string_ids[1], "\\.")[[1]][1])

string_db <- STRINGdb$new(version = "11.5",
                          species = tax_id,
                          score_threshold = 700,
                          input_directory = out_dir)

string_db$plot_network(string_ids)
```

---

## 6. Specialized Plots

### 6.1. UpSet Plot (Intersections)
```{r upset_plot, fig.width=8, fig.height=6}
intersect_list <- list(
  "Up-regulated" = up_genes,
  "Down-regulated" = down_genes
)

p_upset <- upset(fromList(intersect_list),
                 order.by = "freq",
                 mainbar.y.label = "Intersection Size",
                 sets.x.label = "Total Genes",
                 text.scale = c(1.3, 1.3, 1, 1, 1.5, 1.2),
                 point.size = 3.5,
                 line.size = 1)

print(p_upset)
```

### 6.2. Enrichment Map
```{r emap_plots, fig.width=10, fig.height=10}
enrich_go_up_sim <- pairwise_termsim(enrich_go_up)
p_emap <- emapplot(enrich_go_up_sim,
                   showCategory = 25,
                   color = "p.adjust",
                   layout = "nicely") +
  ggtitle("Enrichment Map (Up-regulated)") +
  theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))

p_emap
ggsave(filename = paste0(out_dir, "/emapplot_UP.pdf"), plot = p_emap, width = 10, height = 10)
```

---

## 7. Prepare Data for Advanced Analysis (GSEA)

To run GSEA or upstream regulator analysis, we need the full dataset, not just the filtered significant genes.

```{r prep_advanced_data, warning=FALSE, message=FALSE}
full_res <- read_csv("../results/deseq_results_Control_vs_Treatment.csv") %>%
  mutate(GeneID = str_extract(...1, "WP_[0-9]+\\.[0-9]+")) %>%
  filter(!is.na(GeneID) & !is.na(log2FoldChange))

ranked_list <- full_res %>%
  arrange(desc(log2FoldChange)) %>%
  dplyr::select(GeneID, log2FoldChange)

full_annotation <- full_res %>%
  left_join(eggnog_data, by = c("GeneID" = "X.query"))

saveRDS(ranked_list, file = paste0(adv_dir, "/gsea_ranked_list.rds"))
write_csv(full_annotation, file = paste0(adv_dir, "/full_annotated_results.csv"))
```

---

## Session information
```{r session_info}
sessionInfo()
```
